---
title: "Data 612 Project 5 | Implementing a Recommender System on Spark"
author: "Stephen Haslett"
date: "7/06/2020"
output:
  rmdformats::readthedown

subtitle: "Implementing a Recommender System on Spark"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(recommenderlab)
library(sparklyr)
library(ggplot2)
library(rmdformats)
library(Metrics)
```

## Assignment Instructions
Adapt one of your recommendation systems to work with Apache Spark and compare the performance with your previous iteration.
Consider the efficiency of the system and the added complexity of using Spark.

Please include in your conclusion: For your given recommender system’s data, algorithm(s), and (envisioned) implementation, at
what point would you see moving to a distributed platform such as Spark becoming necessary?

## Introduction
For this assignment, we will create 2 ALS recommender systems based on the MovieLense dataset. One system will be created using the
recommenderlab package (the non-distributed system), and the other will be integrated with Spark using the sparklyr package (the distributed system).

My goal is to then compare the performance of the 2 systems to assess the advantages that distributed systems hold over non-distributed systems.

## Data
Here, we load the version of the MovieLense dataset that comes with the recommenderlab package. The dataset contains around 100,000 ratings
from 943 users on 1664 movies.
```{r, warning=FALSE, message=FALSE}
set.seed(150)
data(MovieLense)
show(MovieLense)
```

After loading the data, we select only users who have rated at least 50 movies, and movies that have been watched at least 100 times.
```{r, warning=FALSE, message=FALSE}
movie_ratings <- MovieLense[rowCounts(MovieLense) > 50, colCounts(MovieLense) > 100]
```


## Non Distributed Recommender System
In this section we will build the non-distributed recommender system using recommenderlab. Here, we split the MovieLense data,
fit the ALS model, perform predictions on the model, and lastly evaluate the accuracy of the model.

#### Build the RecommenderLab ALS based model
```{r, warning=FALSE, message=FALSE}
e <- evaluationScheme(movie_ratings, method = 'split', train = 0.9, given = 15)
rec_als_model <- Recommender(getData(e, 'train'), method = 'ALS')
rec_als_model
```

### Perform predictions on the model and evaluate its accuracy.
```{r, warning=FALSE, message=FALSE}
start <- Sys.time()
predicted <- predict(rec_als_model, getData(e, 'known'), type = 'ratings')
stop <-Sys.time()
rec_als_model_accuracy <- calcPredictionAccuracy(predicted, getData(e, 'unknown'))
rec_als_model_accuracy

rec_time_taken <- stop - start
rec_time_taken
```

## Distributed Recommender System
In this section we will build the distributed recommender system using the sparklyr package to integrate the system with Spark.

Here, we firstly transform the data into a format that is agreeable with Spark, and then connect with Spark so that we can copy the
data over to Spark as an object.

Next we split the data in Spark, and then fit the ALS model. Finally, we make our predictions and monitor
the amount of time it takes Spark to complete processing the data.

#### Build the distributed ALS model and perform predictions
```{r, warning=FALSE, message=FALSE}
# Transform the data.
distributed_data <- movie_ratings %>% as(. , 'data.frame') %>% mutate(user = as.numeric(user), item = as.numeric(item)) 

# Open Spark connection.
sc <- spark_connect(master = "local", version = '3.0.0')

# Copy data to Spark.
spark_data <- sdf_copy_to(sc, distributed_data, 'sdf_rating_matrix', overwrite = TRUE)

# Split the Spark data into testing and training sets and build the model.
split_data <- spark_data %>% sdf_random_split(training = 0.9, testing = 0.2)
distributed_als_model <- ml_als(split_data$training, max_iter = 5)

# perform predictions.
start <- Sys.time()
predicted <- ml_transform(distributed_als_model, split_data$testing) %>% collect()
stop <-Sys.time()
dis_time_taken <- stop - start

# Close Spark connection.
spark_disconnect(sc)
```


## Side by Side Performance Evaluation
To evaluate the performance of the models, we will compare the time it took for each model to complete processing, and also compare the RMSE of both models.

### Processing time comparision

```{r, warning=FALSE, message=FALSE}
processing_time <- rbind(rec_time_taken, dis_time_taken)

rownames(processing_time) <- c('RecommenderLab Model','Distributed Model')
colnames(processing_time) <- c('Processing Time')

knitr::kable(processing_time, format = 'html') %>%
  kableExtra::kable_styling(bootstrap_options = c('striped', 'hover')) %>% 
  add_header_above(c('Model Processing Time Comparision' = 2))
```

### RMSE comparision

```{r, warning=FALSE, message=FALSE}
# Get the RMSE from the RecommenderLab model. RecommenderLabs "calcPredictionAccuracy()" function takes
# care of calculating this for us, so we just need to fetch the value.
rec_rmse <- rec_als_model_accuracy[[1]]

# For the distributed model, we need to calcuate the RMSE ourselves. The Metrics package provides us
# with the rmse() function to take care of this task.
dis_rmse <- rmse(predicted$rating, predicted$prediction)


rmses <- rbind(rec_rmse, dis_rmse)

rownames(rmses) <- c('RecommenderLab Model','Distributed Model')
colnames(rmses) <- c('RMSE')

knitr::kable(rmses, format = 'html') %>%
  kableExtra::kable_styling(bootstrap_options = c('striped', 'hover')) %>% 
  add_header_above(c('Model RMSE Comparison' = 2))
```












