---
title: "Data 612 Project 5 | Implementing a Recommender System on Spark"
author: "Stephen Haslett"
date: "7/06/2020"
output:
  rmdformats::readthedown

subtitle: "Implementing a Recommender System on Spark"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra)
library(recommenderlab)
library(sparklyr)
library(ggplot2)
library(rmdformats)
```

## Assignment Instructions
Adapt one of your recommendation systems to work with Apache Spark and compare the performance with your previous iteration.
Consider the efficiency of the system and the added complexity of using Spark.

Please include in your conclusion: For your given recommender system’s data, algorithm(s), and (envisioned) implementation, at
what point would you see moving to a distributed platform such as Spark becoming necessary?

## Introduction
For this assignment, I adapted the MovieLense recommender system I created for project 2 to work with Apache Spark. I then compared the performance
of the new distrbuted system with your previous project 2 iteration.

### Load the MovieLense Dataset
The MovieLense dataset contains 943 users, and 1664 movies.
```{r, warning=FALSE, message=FALSE}
set.seed(150)
data(MovieLense)
show(MovieLense)
```

## Non Distributed Recommender System

Ratings are integer values ranging from 0 to 5. A zero value rating represents a missing value. As you can see from the below table,
there are a lot of missing values so in the interest of keeping our data accurate, we can remove these.

#### Remove empty rating values from the dataset.

```{r, warning=FALSE, message=FALSE}
ratings_vector <- as.vector(MovieLense@data)
ratings_vector <- ratings_vector[ratings_vector != 0]
ratings_table <- table(ratings_vector)
ratings_table
```

### select only users who have rated at least 50 movies, and movies that have been watched at least 100 times.

```{r, warning=FALSE, message=FALSE}
movie_ratings <- MovieLense[rowCounts(MovieLense) > 50, colCounts(MovieLense) > 100]
```


#### Build the ALS based model
```{r, warning=FALSE, message=FALSE}
e <- evaluationScheme(movie_ratings, method = 'split', train = 0.9, given = 15)
als_model <- Recommender(getData(e, 'train'), method = 'ALS')
als_model
```

### Apply the model to the test data and evaluate the models accuracy.
```{r, warning=FALSE, message=FALSE}
predicted <- predict(als_model, getData(e, 'known'), type = 'ratings')
als_model_accuracy <- calcPredictionAccuracy(predicted, getData(e, 'unknown'))
als_model_accuracy
```

## Distributed Recommender System
```{r, warning=FALSE, message=FALSE}
# Open Spark connection.
sc <- spark_connect(master = "local", version = '3.0.0')

# Close Spark connection.
spark_disconnect(sc)
```
