---
title: "Data 612 Project 4 | Accuracy and Beyond"
author: "Stephen Haslett"
date: "6/28/2020"
output:
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(recommenderlab)
library(ggplot2)
library(rmdformats)
library(rsvd)
library(BBmisc)
```

## Assignment Instructions
1. As in your previous assignments, compare the accuracy of at least two recommender system algorithms against your offline data.

2. Implement support for at least one business or user experience goal such as increased serendipity, novelty, or diversity.

3. Compare and report on any change in accuracy before and after you’ve made the change in #2.

4. As part of your textual conclusion, discuss one or more additional experiments that could be performed and/or metrics that could be
   evaluated only if online evaluation was possible. Also, briefly propose how you would design a reasonable online evaluation environment.

## Introduction
For my previous projects, I used the MovieLense dataset so for this project I will be using the Jester joke rating dataset - http://eigentaste.berkeley.edu/dataset. The recommederlab package ships with a thined down version of the orignal version, so I will be using this version of the dataset. 

## Data Manipulation

### Load the Jester5k dataset
```{r, warning=FALSE, message=FALSE}
set.seed(150)
data(Jester5k)
show(Jester5k)
```


The following image displays the first 25 users' ratings of jokes 1 though 40.  
```{r, warning=FALSE, message=FALSE}
image(Jester5k[1:25, 1:40], "Jester Ratings Sample")
```

The red cells represent positively rated jokes, the blue cells represent negatively rated items, and the white cells represent jokes without ratings.

### Distribution of ratings

```{r, warning=FALSE, message=FALSE}
hist(getRatings(Jester5k),
     main = 'Distribution of Ratings',
     ylab = 'Frequency',
     xlab = 'Rating',
     col = 'Tomato')

```


