---
title: "Data 612 Project 4 | Accuracy and Beyond"
author: "Stephen Haslett"
date: "6/28/2020"
output:
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(recommenderlab)
library(ggplot2)
library(rmdformats)
library(rsvd)
library(BBmisc)
library(tidyverse)

```

## Assignment Instructions
1. As in your previous assignments, compare the accuracy of at least two recommender system algorithms against your offline data.

2. Implement support for at least one business or user experience goal such as increased serendipity, novelty, or diversity.

3. Compare and report on any change in accuracy before and after you’ve made the change in #2.

4. As part of your textual conclusion, discuss one or more additional experiments that could be performed and/or metrics that could be
   evaluated only if online evaluation was possible. Also, briefly propose how you would design a reasonable online evaluation environment.

## Introduction
For my previous projects, I used the MovieLense dataset so for this project, I will be using the Jester joke rating dataset - http://eigentaste.berkeley.edu/dataset. The recommederlab package ships with a thinned down version of the orignal version, so I will be using this version of the dataset. 

## Initial Data Analysis

### Load the Jester5k dataset
```{r, warning=FALSE, message=FALSE}
set.seed(150)
data(Jester5k)
show(Jester5k)
```


The following image displays the first 25 users' ratings of jokes 1 through 40.  
```{r, warning=FALSE, message=FALSE}
image(Jester5k[1:25, 1:40], "Jester Ratings Sample")
```

The red cells represent positively rated jokes, the blue cells represent negatively rated items, and the white cells represent jokes without ratings.

### Distribution of ratings

```{r, warning=FALSE, message=FALSE}
hist(getRatings(Jester5k),
     main = 'Distribution of Ratings',
     ylab = 'Frequency',
     xlab = 'Rating',
     col = 'Tomato')
```

The ratings distribution tells us that the majority of jokes in our subset were given positive ratings. However,
a comparatively small amount of jokes in the set earned the higest rating of 10.

### Average rating across all jokes.
```{r, warning=FALSE, message=FALSE}
mean_rating <- colMeans(Jester5k)
quantile(mean_rating)
```


```{r, warning=FALSE, message=FALSE}
qplot(mean_rating,
     main = 'Distribution of Average Ratings',
     xlab = 'Average Rating')
```

A look at the average ratings across all jokes shows us that the average rating for a good joke is 1.

### Recommender Algorithms

In this section, I am going to explore the performance of the UBCF, and IBCF recommender algorithms. Additionally, in order
to satisfy requirement 2 of the assignment (introducing novelty), I will utilize the RANDOM model.


```{r, warning=FALSE, message=FALSE}
set.seed(1)
# Define an evaluation scheme and pass it to all 3 models.
evaluation <- evaluationScheme(Jester5k, method = 'split', train = 0.8, k = 1, given = 10, goodRating = 1)
user_based_model <- Recommender(getData(evaluation, 'train'), 'UBCF')
item_based_model <- Recommender(getData(evaluation, 'train'), 'IBCF')
random_model <- Recommender(getData(evaluation, 'train'), 'RANDOM')
```



#### Model Prediction Accuracy

Using recommenderlab's prediction functions, we can calculate the accuracy of each model.

```{r, warning=FALSE, message=FALSE}
user_predict <- predict(user_based_model, getData(evaluation, 'known'), type = 'ratings')
item_predict <- predict(item_based_model, getData(evaluation, 'known'), type = 'ratings')
random_predict <- predict(random_model, getData(evaluation, 'known'), type = 'ratings')

results <- rbind(
  'User Model Accuracy' = calcPredictionAccuracy(user_predict, getData(evaluation, 'unknown')), 
  'Item Model Accuracy' = calcPredictionAccuracy(item_predict, getData(evaluation, 'unknown')),
  'Random Model Accuracy' = calcPredictionAccuracy(random_predict, getData(evaluation, 'unknown'))
)
knitr::kable(results, format = 'html') %>%
  kableExtra::kable_styling(bootstrap_options = c('striped', 'hover'))

```


As reflected in the above table, The least accurate model is the RANDOM model, were as the most accurate model is the UBCF model. This is not surpising
as recommendations based on previous ratings will be more accurate than those picked randomly.

```{r, warning=FALSE, message=FALSE}
models <- list('User Based Model' = list(name = 'UBCF', param = list(nn = 50)),
               'Item Based Model' = list(name = 'IBCF', param = list(k = 50)),
               'Random Model' = list(name = 'RANDOM', param = NULL))

results <- evaluate(evaluation, models, type = 'topNList', n = c(1, 5, 10, 20, 30, 50))
results
```

### Analysis
Next we can explore the Precision Recall and ROC Curve for all 3 models. Again, the UBCF model out performs the other models in terms of
both precision, and ROC.

#### Precision Recall
```{r, warning=FALSE, message=FALSE}
plot(results, 'prec/rec', annotate = 1)
title('Precision Recall')
```



#### ROC Curve
```{r, warning=FALSE, message=FALSE}
plot(results, 'ROC', annotate = 1)
title('ROC Curve')
```

### Introducing Novelty

To introduce novelty into the equation, we will create a hybrid recommendation model that combines the RANDOM model with the top
performing UBCF model. We will then measure the accuracy of the model. 

```{r, warning=FALSE, message=FALSE}
# Create the hybrid model.
hybrid_model <- HybridRecommender(user_based_model, random_model, weights = c(0.8, 0.2))

# Predict the accuracy of the model.
hybrid_predict <- predict(hybrid_model, getData(evaluation, 'known'), type = 'ratings')

hybrid_results <- calcPredictionAccuracy(hybrid_predict, getData(evaluation, 'unknown')) 
knitr::kable(hybrid_results, format = 'html') %>%
  kableExtra::kable_styling(bootstrap_options = c('striped', 'hover'))
```

### Model RMSE Comparision

Finally, we will compare the RMSE of the original UBCF model with that of the new hybrid model so we can get a side by side comparison.

```{r, warning=FALSE, message=FALSE}
user_results <- calcPredictionAccuracy(user_predict, getData(evaluation, 'unknown'))
comparison <- rbind(
  'User Model RMSE' = user_results['RMSE'], 
  'Hybrid Model RMSE' = hybrid_results['RMSE']
)

knitr::kable(comparison, format = 'html') %>%
  kableExtra::kable_styling(bootstrap_options = c('striped', 'hover'))
```

As reflected in the above table, the RMSE is higher for the hybrid model than for the UBCF model and thus UBCF is the most accurate model.

