---
title: "Data 612 Project 3 | Matrix Factorization Methods"
author: "Stephen Haslett"
date: "6/21/2020"
output:
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(recommenderlab)
library(ggplot2)
library(rmdformats)
library(rsvd)
library(BBmisc)
```

## Assignment Instructions
Your task is to implement a matrix factorization method—such as singular value decomposition (SVD) or
Alternating Least Squares (ALS)—in the context of a recommender system.

You may approach this assignment in a number of ways. You are welcome to start with an existing recommender
system written by yourself or someone else. Remember as always to cite your sources,
so that you can be graded on what you added, not what you found.

SVD can be thought of as a pre-processing step for feature engineering. You might easily start with
thousands or millions of items, and use SVD to create a much smaller set of “k” items (e.g. 20 or 70).


## Introduction
For this assignment, I am going to expand upon the MovieLense recommender system I built for Project 2.

## Data Manipulation

### Load the MovieLense dataset
```{r, warning=FALSE, message=FALSE}
set.seed(150)
data(MovieLense)
show(MovieLense)
```


### Select only users who have rated at least 50 movies, and movies that have been watched at least 100 times
```{r, warning=FALSE, message=FALSE}
movie_ratings <- MovieLense[rowCounts(MovieLense) > 50, colCounts(MovieLense) > 100]
movie_ratings <- as.matrix(movie_ratings@data)
```


### Convert 0 values (missing data) to NA values, and then replace the NULL values with row mean values
```{r, warning=FALSE, message=FALSE}
# Impute the data
movie_ratings[movie_ratings == 0] <- NA
value <- which(is.na(movie_ratings), arr.ind = TRUE)
movie_ratings[value] <- rowMeans(movie_ratings, na.rm = TRUE)[value[,1]]
```


### Split the data into training and test sets.
```{r, warning=FALSE, message=FALSE}
train <- sample(x = c(TRUE, FALSE), size = nrow(movie_ratings), replace = TRUE, prob = c(0.8, 0.2))
training_data <- movie_ratings[train, ]
test_data <- movie_ratings[!train, ]

print(nrow(training_data))
print(nrow(test_data))
```


### Normalize the data and apply the SVD model
```{r, warning=FALSE, message=FALSE}
# Normalize the data and then apply the SVD model.
normalized_data <- normalize(movie_ratings)
svd_data <- svd(normalized_data)
```


## Model Evaluation
```{r, warning=FALSE, message=FALSE}
movie_data <- as(normalized_data, 'realRatingMatrix')
e <- evaluationScheme(movie_data, method = 'split', train = 0.9, given = 15, goodRating = 3, k = 10)
ubcf <- Recommender(getData(e, 'train'), 'UBCF')
ibcf <- Recommender(getData(e, 'train'), 'IBCF')
svd_model <- Recommender(getData(e, 'train'), 'SVD')

predict_ubcf <- predict(ubcf, getData(e, 'known'), type = 'ratings')
predict_ibcf <- predict(ibcf, getData(e, 'known'), type = 'ratings')
predict_svd <- predict(svd_model, getData(e, 'known'), type = 'ratings')

ubcf_error <- calcPredictionAccuracy(predict_ubcf, getData(e, 'unknown'))
ibcf_error <- calcPredictionAccuracy(predict_ibcf, getData(e, 'unknown'))
svd_error <- calcPredictionAccuracy(predict_svd, getData(e, 'unknown'))

error <- rbind(ubcf_error, ibcf_error, svd_error)
rownames(error) <- c('UBCF','IBCF', 'SVD')

error
```

### Prediction Error Comparision
```{r, warning=FALSE, message=FALSE}
barplot(error,
        main = "Prediction Error Comparison",
        ylab = 'Error Rate',
        xlab = 'Error Type',
        col=c('RoyalBlue','Tomato', 'YellowGreen'),
        legend = rownames(error),
        beside = TRUE)
```
